Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-12 paper.tex
Edites by yan

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
SARS-CoV2 shows little recombination, with recombination nodes accounting
for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be adequately represented by a
single tree, particularly early in the pandemic, or when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives.

For validation purposes, we compare the genetic
relationships in our ARGs with phylogenetic relationships
inferred by the NextStrain project \citep{Hadfield2018-ef}.
To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
the sc2ts ARGs to a ``backbone'' containing only samples present
in a specific NextStrain tree. For this purpose, we picked an
``all-data''  NextStrain tree restricted to the time period covered
by the appropriate ARG. This results in a relatively small
set of shared samples, almost entirely restricted to non-recombinant lineages.
These simplified backbone ARGs exhibit very
little recombination (6 and 5 trees across the genome for the Wide and Long
ARGs, respectively).
% JK: This leaves a big question hanging - why do we ahave recombination
% in here at all? It's probably just artefactual, right?

>>>>

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the NextStrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded ``all-data'' NextStrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the sc2ts ARGs to a ``backbone'' containing only samples present
the NextStrain tree. This results in a small set of shared
samples (180 for the wide ARG, 88 for the long ARG), none of which are
designated as Pango recombinant lineages by Nextclade (see subsection
\ref{sec:pango_x_lineages}).

These sc2ts backbone phylogenies do contain small amounts of recombination
